---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

  * Load the data
```{r, echo = FALSE, message = FALSE}
    library(data.table) 
    library(lubridate)
```
```{r, echo = TRUE}
    activity <- data.table(read.csv("activity.csv", sep = ",", header = TRUE))
```

## What is mean total number of steps taken per day?
  * Total number of steps taken per day
```{r}
    activity[, totalActivity := sum(steps), by = date]
```
 
```{r, echo = FALSE}
    hist(activity$totalActivity, breaks=10, main = "Histogram of the total number of steps taken each day", xlab = "Total steps per day" ) 
```
 
 * Mean of total number of steps taken per day
```{r, echo = FALSE}
    mean(activity$totalActivity, na.rm = TRUE)
```
 
 * Median  of total number of steps taken per day
```{r, echo = FALSE}
    median(activity$totalActivity, na.rm = TRUE)
```

## What is the average daily activity pattern?
* Average of steps per 5-minute interval
```{r}
    activityPerInterval <- activity[, list(averageActivity = mean(steps, na.rm = TRUE)), by = interval]
```
```{r, echo = FALSE}
    plot(activityPerInterval, type="l", main = "Average daily activity pattern", xlab = "Interval", ylab = "Average steps", col = "red")
```
 
 * Maximum average per 5-minulte interval
```{r, echo = FALSE}
    max(activityPerInterval$averageActivity)
```
 
## Imputing missing values
* Total number of missing values in the data set
```{r, echo = FALSE}
    activity[, sum(is.na(steps))]
```
 
 * New data set with the missing data filled in using the average per 5-minute interval across all days.
```{r, echo = TRUE, warning = FALSE}
 
    filledActivity <- activity[, list(date, filledSteps = mean(steps, na.rm = TRUE), steps), by = interval][order(date, interval)][is.na(steps), steps := filledSteps]

```

 * Total number of steps taken per day
```{r}
    filledActivity[, totalActivity := sum(steps), by = date]
```
 
```{r, echo = FALSE}
    par(mfrow = c(2,1), mar = c(4, 4, 3, 2))
    hist(activity$totalActivity, breaks=20, main = "Histogram of the total number of steps taken each day", xlab ="")
    hist(filledActivity$totalActivity, breaks=20, main = "Histogram of the total number of steps taken each day (NA s removed)", xlab = "Total steps per day")
```

 * Mean of total number of steps taken per day
```{r, echo = FALSE}
    mean(activity$totalActivity, na.rm = TRUE)
```
 
 * Median of total number of steps taken per day
```{r, echo = FALSE}
    median(activity$totalActivity, na.rm = TRUE)
```
 
 * Mean of total number of steps taken per day (NA s removed)
```{r, echo = FALSE}
    mean(filledActivity$totalActivity, na.rm = TRUE)
```
 
 * Median of total number of steps taken per day (NA s removed)
```{r, echo = FALSE}
    median(filledActivity$totalActivity, na.rm = TRUE)
```
 
 
## Are there differences in activity patterns between weekdays and weekends?

```{r, echo = FALSE}
    weekend = c(1, 7)
    weekdays = c(2, 3, 4, 5, 6)
```
 * Add new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r, echo = TRUE}
    filledActivity[wday(as.Date(date)) %in% weekend, day := "weekend"]
    filledActivity[wday(as.Date(date)) %in% weekdays, day := "weekday"]
    filledActivity[,day := factor(day)]
    
    activityPerIntervalWeekends <- filledActivity[day == "weekend", list(averageActivity = mean(steps, na.rm = TRUE)), by = interval]
    activityPerIntervalWeekdays <- filledActivity[day == "weekday", list(averageActivity = mean(steps, na.rm = TRUE)), by = interval]
```
```{r, echo = FALSE}    
    par(mfrow = c(2,1), mar = c(4, 4, 3, 2))
    
    plot(activityPerIntervalWeekends, type="l", main = "Average daily activity pattern Weekends", col = "red")
    plot(activityPerIntervalWeekdays, type="l", main = "Average daily activity pattern Weekdays", col = "red")
```